\sffamily An EEG-based visual Brain-Computer Interface for gaze-free communication

### 2024-10-09 chapters/patients.tex
Comments François patients

Eye motor impairments are presumed to reduce performance in operating visual
oddball \ac{bci}s (see \cref{sec:gaze-independence} for an overview), since users
cannot comfortably redirect their gaze at the desired target,
i.e., perform overt \ac{vsa}.
This is usually circumvented by designing gaze-independent~
\acp{bci}~\cite{Riccio2012}.
These interfaces either avoid visual stimulation or exploit some form of
covert \ac{vsa}, where the gaze and \ac{vsa} do not coincide.

@@

It is a striking constatation that studies reporting on
gaze-independent visual \ac{bci} with people with \ac{sspi} and eye-motor
impaired are very few.
Results are usually different than those obtained with healthy control
participants in the lab, due to difference in capabilities, brain
response, equipment and environment.

\textcite{Lesenfants2014} tested a gaze-independent \ac{ssvep} \ac{bci} in six
participants with \ac{lis} yet only exceeded chance level accuracy in two.
More recently, \textcite{Peters2020} performed a trial with two
participants with late-stage \ac{als} and visual impairment.
Their \ac{ssvep} interface was not optimized for gaze-independence, but the
system showed high accuracy, outperforming an eye tracking alternative.
It would be of interest to verify if such results can be replicated with other
patient groups and visual oddball \acp{bci}.

@@

Our previous study, presented in \cref{sec:covert-align} also used the
visual Hex-o-Spell interface~\cite{VanDenKerchove2024}.
This work partially accounted for the idea that \ac{bci} users with \ac{sspgi} might
not fully rely on central gaze fixation and evaluated settings that are not
strictly dependent on this.
We showed gaze-independent performance can be improved in healthy subjects
using a suited decoding strategy.
That accounts for latency jitter in covert \ac{vsa} responses.
Yet, there is a strong need for verification of these results in an applied
setting with people with \ac{sspi}.

@@

Ultimately, 7 participants were retained.
Of these, one participant was diagnosed with bulbar-onset \ac{als}.
\ac{als} is a neurodegenerative disease affecting the motor neurons, leading to
progressive loss of motor function.
This initially results in general weakness and loss of muscle tone, but
eventually leads to full body paralysis.
Although speech and especially eye movements are usually preserved until the
later stages of the disease progression, the bulbar-onset variant is
characterized by an early loss of speech and an increased involvement of eye
motor symptoms~\cite{Guo2022}.
Furthermore, one of the goals of \ac{bci} is to support individuals with \ac{als}
whose life span has been extended with life support and assistive technology to
ensure quality of life.
In these very progressed stages of \ac{als}, eye movement will eventually also
be affected~\cite{Hayashi1991}.

Three other participants were diagnosed with \ac{fa}, a neurodegenerative
disease affecting the
spinal cord, peripheral nervous system, and cerebellum.
\ac{fa} results in an impairing loss of muscle coordination.
In progressed stages, the disease can present with nystagmus, saccadic
intrusions, and gaze dysmetria~\cite{Cook2017}.

The final three participants were recovering from stroke.
Brain stem and cerebellar stroke can often lead to forms of \ac{lis}, and
various eye motor disorders are common depending on the exact etiology
~\cite{Bogousslavsky1987, Moncayo2009}.
Stroke can lead to some of the more severe impairments like ophthalmoplegia or
partial ophthalmoplegia immediately from onset.

Table~\ref{tab:patients/patients} lists the included participants and their
diagnoses.

@@

Self-reported eye motor and visual abnormalities were recorded according to the
relevant visual \ac{bci} skills presented by~\textcite{FriedOken2020}.
These include visual acuity, visual fixation, eyelid function, ocular motility,
binocular vision, and field of vision.
Additionally, participants were asked about eye tremors (nystagmus or other) and
other involuntary eye movements.

@@

Table~\ref{tab:patients/eye} details the eye motor impairments and vision of
the included participants.
All participants reported some degree of fatigue or discomfort when fixating.
Participant PA1 had the mildest impairment, only reporting fatigue when fixating
for prolonged times.
The \ac{fa} participants were mostly affected by eye tremors and impaired pursuit.
PB2 suffered from especially severe horizontal oscillating involuntary eye
movements.
Eye motor function of participants PC2, PC3, and PC4 was most severely affected.
Participant PC2 was only able to look up and down and had a deviation in the
left eye causing diplopia, but this was corrected by a prism glass.
Participant PC3 only retained partial motility of the right eye, while the left eye was permanently closed.
Participant PC4 had one deviated eye with a corneal abscess affecting the motility
and vision in the right eye, and suffered reduced motility in the left.

\begin{table}[t]
  \centering
  \footnotesize
  \input{chapters/patients_table_eye.tex}
  \caption[Visual skills of the included participants.]{%
  Visual skills of the included participants.
  Visual~\ac{bci} skills~\cite{FriedOken2020}
  were assessed with a combination of self-reported issues
  by the subject and the NeuroEye~\cite{Hassan2022} test. \BigCross\
  impaired, \BigDiamondshape\ severely impaired.
  logMAR: lower is better.}
  \label{tab:patients/eye}
\end{table}

@@

\fullpagefig{%
  \includegraphics[width=\textwidth]{figures/patients/fig_gaze.png}
}{%
  Distribution of the recorded gaze position.
}{%
  Distribution of the recorded gaze position during the experimental session in the three \ac{vsa}
  conditions.
  Crosshairs represent stimulus positions, with the orange ones cued during
  the given condition.
  Subjects PB2 and PC4 preferred covert \ac{bci} operation, with PB2 resting gaze
  near the middle of the screen, and PC4 near the bottom.
}{%
  fig:patients/gaze
}

PA1 had relatively intact gaze control and was able to correctly perform the
cued overt and covert settings.
When gaze was uncued, he fixated on the cued target.
This was also mostly the case for PB1, although eye tracking revealed that he
chose not to perform central gaze fixation when cued in at least one of the
stimulation blocks. We were unable to record his gaze near the bottom-left
stimulus position, either due to eye tracker failure or because the participant
was not comfortable fixating on this position.
Eye tracker calibration did not succeed for subject PB4, but given
transformations of gaze positions to the stimulus space, they were assumed to be overtly performing the task.

>>>>

Eye motor impairments are presumed to reduce performance in operating visual
oddball \ac{bci}s (see \cref{sec:gaze-independent} for an overview), since users
cannot comfortably redirect their gaze at the desired target,
i.e., perform overt \ac{vsa}.
This is usually circumvented by designing gaze-independent~
\acp{bci}~\cite{Riccio2012}.
These interfaces either avoid visual stimulation or exploit some form of
covert \ac{vsa}, where the gaze and \ac{vsa} do not coincide.

@@

It is a striking constatation that studies reporting on
gaze-independent visual \ac{bci} with people with \ac{sspi} and eye-motor
impaired are very few.
Results are usually different from those obtained with healthy control
participants in the lab, due to difference in capabilities, brain
response, equipment and environment.

\textcite{Lesenfants2014} tested a gaze-independent \ac{ssvep} \ac{bci} in six
participants with \ac{lis} yet only exceeded chance level accuracy in two.
More recently, \textcite{Peters2020} performed a trial with two
participants with late-stage \ac{als} and visual impairment.
Their \ac{ssvep} interface was not optimized for gaze-independence, but the
system showed high accuracy, outperforming an eye tracking alternative.
It would be of interest to verify if such results can be replicated with
participants with other conditions, and with a visual oddball \acp{bci}.

@@

Our previous study, presented in \cref{sec:covert-align} also used the
visual Hex-o-Spell interface~\cite{VanDenKerchove2024}.
This work partially accounted for the idea that \ac{bci} users with \ac{sspgi} might
not fully rely on central gaze fixation and evaluated settings that are not
strictly dependent on this.
We showed gaze-independent performance can be improved in healthy subjects
using a suited decoding strategy that accounts for latency jitter in covert \ac{vsa} responses.
Yet, there is a strong need for verification of these results in an applied
setting with people with \ac{sspi}.

@@

Ultimately, 7 participants were retained.
Of these, one participant was diagnosed with bulbar-onset \ac{als}, three with
\ac{fa} an three were recovering from stroke.
We refer to \cref{sec:gaze-independent/gaze-dependence/incidence} for a short
description of these conditions.
Table~\ref{tab:patients/patients} lists the included participants and their
diagnoses.

@@

Self-reported eye motor and visual abnormalities were recorded according to the
relevant visual \ac{bci} skills presented by~\textcite{FriedOken2020}.
These include visual acuity, visual fixation, eyelid function, ocular motility,
binocular vision, and field of vision.
Additionally, participants and their caregivers were asked about eye tremors
(nystagmus or other) and other involuntary eye movements.

@@

Table~\ref{tab:patients/eye} details the eye motor impairments and vision of
the included participants.
All participants reported some degree of fatigue or discomfort when fixating.
Participant PA1 had the mildest impairment, only reporting fatigue when fixating
for prolonged times.
The \ac{fa} participants were mostly affected by eye tremors and impaired pursuit.
PB2 suffered from especially severe horizontal oscillating involuntary eye
movements.
Eye motor function of participants PC2, PC3, and PC4 was most severely affected.
Participant PC2 was only able to look up and down and had a deviation in the
left eye causing diplopia, but this was corrected by a prism glass.
Participant PC3 only retained partial motility of the right eye, while the left eye was permanently closed.
Participant PC4 had one deviated eye with a corneal abscess affecting the motility
and vision in the right eye, and reducing motility in the left.

\begin{table}[t]
  \centering
  \footnotesize
  \input{chapters/patients_table_eye.tex}
  \caption[Visual skills of the included participants.]{%
  Visual skills of the included participants.
  Visual~\ac{bci} skills~\cite{FriedOken2020}
  were assessed with a combination of self-reported issues
  by the subject and the NeuroEye~\cite{Hassan2022} test. \BigCross\
  impaired, \BigDiamondshape\ severely impaired.
  logMAR: lower is better.}
  \label{tab:patients/eye}
\end{table}
\todo{where these values corrected for distance?}

@@

\fullpagefig{%
  \includegraphics[width=\textwidth]{figures/patients/fig_gaze.png}
}{%
  Distribution of the recorded gaze position.
}{%
  Distribution of the recorded gaze position during the experimental session in the three \ac{vsa}
  conditions.
  Crosshairs represent stimulus positions, with the orange ones cued during
  the given condition.
  Subjects PB2 and PC4 preferred covert \ac{bci} operation, with PB2 resting gaze
  near the middle of the screen, and PC4 near the bottom.
}{fig:patients/gaze}

PA1 had relatively intact gaze control and was able to correctly perform the
cued overt and covert settings.
When gaze was uncued, he fixated on the cued target.
This was also mostly the case for PB1, although eye tracking revealed that he
chose not to perform central gaze fixation when cued in at least one of the
stimulation blocks. We were unable to record his gaze near the bottom-left
stimulus position, either due to eye tracker failure or because the participant
was not comfortable fixating on this position.
Eye tracker calibration did not succeed for subject PB4, but given
transformation of gaze positions to the stimulus space, they were assumed to
be overtly performing the free task.
